% !TeX spellcheck = en_GB

\documentclass[thesis=M,english,hidelinks]{FITthesis}[2018/03/12]

\usepackage[utf8]{inputenc}

\usepackage{hyperref}

\usepackage{graphicx}
\graphicspath{{./figures/}}

\usepackage{dirtree}

\usepackage{amsmath} % maths
\DeclareMathOperator{\arctantwo}{arctan2}
\DeclareMathOperator{\sgn}{sgn}

\usepackage[]{algorithm2e}

\usepackage{listings}

\department{Department of Theoretical Computer Science}
\title{Exploring use of non-negative matrix factorization for lossy audio compression}
\authorGN{Tomáš} %author's given name/names
\authorFN{Drbota} %author's surname
\author{Tomáš Drbota} %author's name without academic degrees
\authorWithDegrees{Bc. Tomáš Drbota} %author's name with academic degrees
\supervisor{doc. Ing. Ivan Šimeček, Ph.D.}
\acknowledgements{I would like to thank my supervisor for always being quick to consult any arising issues. My thanks also go to my family and friends for supporting me throughout the entire time of my studies.}
\abstractEN{Non-negative matrix factorization has been successfully applied in various scenarios involving analysis of large chunks of data and finding patterns in them for later use. It's used to perform things such as face recognition, source separation or image compression among others.

The purpose of this thesis is to research possible uses of non-negative matrix factorization in the problem of lossy audio compression. A reference audio encoder and decoder using NMF will be implemented and various experiments using this encoder will be conducted. The results will be measured and compared to existing audio compressing solutions.}
\abstractCS{Algoritmus nezáporného rozkladu matic byl úspěsně použit pro mnoho různých problémů, které souvisí s analýzou velkého množství dat a z nich plynoucích souvislostí. Využívá se např. pro rozpoznávání obličejů, separaci signálů či kompresi obrázků.

Cílem této práce je prozkoumat možné využití nezáporného rozkladu matic pro ztrátovou kompresi zvuku. Bude naprogramován referenční kodér a dekodér zvuku využívajicí NMF, a s ním budou potom provedeny různé experimenty. Výsledky budou změřeny a srovnány vůči existujícím nástrojům pro kompresi zvuku.}
\placeForDeclarationOfAuthenticity{Prague}
\keywordsCS{ztrátová komprese, zvuk, komprese dat, nmf, kódování}
\keywordsEN{lossy compression, audio, data compression, nmf, encoding}
\declarationOfAuthenticityOption{5}
% \website{http://site.example/thesis} %optional thesis URL

\begin{document}
	
\addtocontents{toc}{\setcounter{tocdepth}{2}}

\chapter{Introduction}
\input{chapters/introduction}

\chapter{Digital audio}
\input{chapters/digital_audio}

\chapter{Non-negative matrix factorization}
\input{chapters/nmf}

\chapter{Data compression}
\input{chapters/data_compression}

\chapter{Design}
\input{chapters/design}

\chapter{Implementation}
\input{chapters/implementation}

\chapter{Evaluation}
\input{chapters/evaluation}

\chapter{Conclusion}
\input{chapters/conclusion}

\bibliographystyle{iso690}
\bibliography{mybibliographyfile}

\appendix

\chapter{List of Symbols}
\begin{tabular}{r l}
	$t$ & symbol representing a time value in seconds \\
	$\tau$ & symbol representing a "slow" time, time index with a lower resolution than $t$ \\
	$\xi$ & symbol representing a frequency value in hertz \\
	$F_s$ & symbol representing a sampling rate of an audio signal \\
	$z$ & symbol representing a complex number \\
	$\phi(z)$ & symbol representing the phase of a complex number \\
	$|z|$ & symbol representing the magnitude of a complex number \\
	$Q(x)$ & function representing a uniform quantizer \\
	$\Delta$ & symbol representing the step size of a uniform quantizer \\
	$F(x)$ & function representing $\mu$-law compression \\
	$x(t)$ & function representing the amplitude of a continuous signal at a time $t$ \\
	$x_n$ & sequence representing the amplitude of a discrete signal indexed by $n$ \\
	$w(t)$ & continuous windowing function at a time $t$ \\
	$w_n$ & discrete windowing function indexed by $n$ \\
	$X(\xi)$ & function representing the frequency component of a signal for a frequency $\xi$ \\
	$S(\xi)$ & the Fourier transform of a continuous signal \\
	$S_k$ & the discrete Fourier transform of a discrete signal \\
	$S(\tau, \xi)$ & the short-time Fourier transform of a continuous signal \\
	$S_{k, \xi}$ & the discrete short-time Fourier transform of a discrete signal \\
	$M_k$ & the Modified discrete cosine transform of a discrete signal \\
	$r$ & rank of a NMF approximation \\
	$S_f$ & frame size \\
	$S_c$ & chunk size
\end{tabular}

\chapter{Abbreviations}
\begin{description}
	\item[NMF] non-negative matrix factorization
	\item[DFT] discrete Fourier transform
	\item[STFT] short-time Fourier transform
	\item[FFT] Fast Fourier Transform
	\item[DCT] discrete cosine transform
	\item[MDCT] modified discrete cosine transform
\end{description}


\chapter{Contents of enclosed CD}

\begin{figure}
	\dirtree{%
		.1 README.txt\DTcomment{the file with CD contents description}.
		.1 src\DTcomment{the directory of source codes}.
		.2 audionmf\DTcomment{implementation sources}.
		.2 thesis\DTcomment{the directory of \LaTeX{} source codes of the thesis}.
		.1 text\DTcomment{the thesis text directory}.
		.2 thesis.pdf\DTcomment{the thesis text in PDF format}.
	}
\end{figure}

\end{document}
